SOBP_Analyses_BHRC_ABCD_ACEsCog

### 2021-11-17 SOBP_Analyses_BHRC_ABCD_ACEsCog.Rmd
final analyses for abstract
this includes each of the analyses for the SOBP abstract

# Calculate Thompson factor scores, remove outliers, calculate total ACEs
```{r}
combined.data$thompson_PC1 <- (0.754*combined.data$PicVocab)+(0.213*combined.data$Flanker)+(0.471*combined.data$List)+(0.205*combined.data$CardSort)+(0.015*combined.data$Pattern)+(0.012*combined.data$Picture)+(0.82*combined.data$Reading)+(0.306*combined.data$RAVLT)+(0.5*combined.data$LMT)

@@

combined.data.noOutliers <- combined.data[(combined.data$thompson_PC1>(mean_PC1-sd3_PC1) & combined.data$thompson_PC1<(mean_PC1+sd3_PC1) & combined.data$thompson_PC2>(mean_PC2-sd3_PC2) & combined.data$thompson_PC2<(mean_PC2+sd3_PC2) & combined.data$thompson_PC3>(mean_PC3-sd3_PC3) & combined.data$thompson_PC3<(mean_PC3+sd3_PC3)),]

combined.data.noOutliers[,224:240]<-as.numeric(unlist(combined.data.noOutliers[,224:240]))

combined.data.noOutliers$abuse <- rowSums(data.frame(as.numeric(combined.data.noOutliers$ksads_ptsd_raw_761_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_762_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_763_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_767_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_768_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_769_p)),na.rm=TRUE)
combined.data.noOutliers$totalACEs <- rowSums(data.frame(as.numeric(combined.data.noOutliers$ksads_ptsd_raw_754_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_755_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_756_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_757_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_758_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_759_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_760_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_761_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_762_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_763_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_764_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_765_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_766_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_767_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_768_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_769_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_770_p)),na.rm=TRUE)

family <-read.table("/Users/ariellekeller/Documents/Kellernet_PrelimAnalysis/ses/acspsw03.txt",header=TRUE)
family.baseline<-family[family$eventname=="baseline_year_1_arm_1",]
abcd.data <- merge(combined.data.noOutliers,family.baseline[, c("subjectkey", setdiff(colnames(family.baseline),colnames(combined.data.noOutliers)))],by="subjectkey")

@@

# https://www.sciencedirect.com/science/article/pii/S2352289518300821
abcd.data$acceptance <- rowSums(data.frame(as.numeric(abcd.data$crpbi_parent1_y),as.numeric(abcd.data$crpbi_parent2_y),as.numeric(abcd.data$crpbi_parent3_y),as.numeric(abcd.data$crpbi_parent4_y),as.numeric(abcd.data$crpbi_parent5_y)))

abcd.data$monitoring <- rowSums(data.frame(as.numeric(abcd.data$parent_monitor_q1_y),as.numeric(abcd.data$parent_monitor_q2_y),as.numeric(abcd.data$parent_monitor_q3_y),as.numeric(abcd.data$parent_monitor_q4_y),as.numeric(abcd.data$parent_monitor_q5_y)))

@@

```{r}
summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+totalACEs+totalACEs*sex,data=abcd.data))
summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+totalACEs+totalACEs*sex,data=abcd.data))
summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+totalACEs+totalACEs*sex,data=abcd.data))

abcd.data$ACEs.binary <- factor(abcd.data$totalACEs>0,labels = c("More Threat","Less Threat"))
abcd.data$ACEs.binary.numeric <- as.numeric(abcd.data$ACEs.binary)

summary(lmer(ACEs.binary.numeric~interview_age+sex+(1|rel_family_id)+thompson_PC1+thompson_PC2+thompson_PC3,data=abcd.data))

summary(lmer(acceptance~interview_age+sex+(1|rel_family_id)+thompson_PC1+thompson_PC2+thompson_PC3,data=abcd.data))
summary(lmer(monitoring~interview_age+sex+(1|rel_family_id)+thompson_PC1+thompson_PC2+thompson_PC3,data=abcd.data))

summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+ACEs.binary.numeric+acceptance+monitoring,data=abcd.data))

summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+ACEs.binary.numeric+acceptance+monitoring,data=abcd.data))

summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+ACEs.binary.numeric+acceptance+monitoring,data=abcd.data))

ggplot(data=abcd.data,aes(totalACEs,thompson_PC1))+geom_point(size=1,colour=abcd.data$interview_age)+geom_smooth(method=lm)+ facet_wrap(~ factor(sex,labels=c("Male","Female")))
```

@@

df.summary <- abcd.data %>%
     group_by(ACEs.binary,agesRounded) %>%
     summarise(
         se = sd(thompson_PC1, na.rm = TRUE)/sqrt(length(thompson_PC1)),
         thompson_PC1 = mean(thompson_PC1))
ggplot(data=df.summary, aes(x=ACEs.binary,y=thompson_PC1)) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("ACEs") + ylab("thompson_PC1") + geom_errorbar(aes(ymin=thompson_PC1-se, ymax=thompson_PC1+se), width=.2,position=position_dodge(.9))+ facet_wrap(~ factor(round(agesRounded)))+coord_cartesian(ylim=c(210,255))
```

>>>>

# Calculate Thompson factor scores and remove outliers
```{r}
combined.data$thompson_PC1 <- (0.754*combined.data$PicVocab)+(0.213*combined.data$Flanker)+(0.471*combined.data$List)+(0.205*combined.data$CardSort)+(0.015*combined.data$Pattern)+(0.012*combined.data$Picture)+(0.82*combined.data$Reading)+(0.306*combined.data$RAVLT)+(0.5*combined.data$LMT)

@@

combined.data.noOutliers <- combined.data[(combined.data$thompson_PC1>(mean_PC1-sd3_PC1) & combined.data$thompson_PC1<(mean_PC1+sd3_PC1) & combined.data$thompson_PC2>(mean_PC2-sd3_PC2) & combined.data$thompson_PC2<(mean_PC2+sd3_PC2) & combined.data$thompson_PC3>(mean_PC3-sd3_PC3) & combined.data$thompson_PC3<(mean_PC3+sd3_PC3)),]
```

## Calculate types of ACEs
```{r}
combined.data.noOutliers[,224:240]<-as.numeric(unlist(combined.data.noOutliers[,224:240]))

# Calculation of subscore with items related to physical and sexual abuse
combined.data.noOutliers$abuse <- rowSums(data.frame(as.numeric(combined.data.noOutliers$ksads_ptsd_raw_761_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_762_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_763_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_767_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_768_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_769_p)),na.rm=TRUE)

# Calculation of subscore with items related to trauma (car accidents, natural disasters, loss, etc.)
combined.data.noOutliers$trauma <- rowSums(data.frame(as.numeric(combined.data.noOutliers$ksads_ptsd_raw_754_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_755_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_756_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_757_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_758_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_759_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_760_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_766_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_770_p)),na.rm=TRUE)

# Calculation of total score from the KSADS items
combined.data.noOutliers$totalACEs <- rowSums(data.frame(as.numeric(combined.data.noOutliers$ksads_ptsd_raw_754_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_755_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_756_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_757_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_758_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_759_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_760_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_761_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_762_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_763_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_764_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_765_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_766_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_767_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_768_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_769_p),as.numeric(combined.data.noOutliers$ksads_ptsd_raw_770_p)),na.rm=TRUE)

# Remember to add in this variable about family structure so we can use it as a covariate
family <-read.table("/Users/ariellekeller/Documents/Kellernet_PrelimAnalysis/ses/acspsw03.txt",header=TRUE)
family.baseline<-family[family$eventname=="baseline_year_1_arm_1",]
abcd.data <- merge(combined.data.noOutliers,family.baseline[, c("subjectkey", setdiff(colnames(family.baseline),colnames(combined.data.noOutliers)))],by="subjectkey")

@@

# Calculate subscales of deprivation (emotional and physical neglect) - see this ref for motivation https://www.sciencedirect.com/science/article/pii/S2352289518300821
abcd.data$acceptance <- rowSums(data.frame(as.numeric(abcd.data$crpbi_parent1_y),as.numeric(abcd.data$crpbi_parent2_y),as.numeric(abcd.data$crpbi_parent3_y),as.numeric(abcd.data$crpbi_parent4_y),as.numeric(abcd.data$crpbi_parent5_y)))

abcd.data$monitoring <- rowSums(data.frame(as.numeric(abcd.data$parent_monitor_q1_y),as.numeric(abcd.data$parent_monitor_q2_y),as.numeric(abcd.data$parent_monitor_q3_y),as.numeric(abcd.data$parent_monitor_q4_y),as.numeric(abcd.data$parent_monitor_q5_y)))
```

## Binarize the four ACEs subtypes (2 "Threat" types: Abuse and Trauma; 2 "Deprivation" types: Emotional Neglect and Physical Neglect)

```{r}
abcd.data$abuse.binary <- factor(abcd.data$abuse>=1,labels = c("No Abuse","Abuse"))
abcd.data$abuse.binary.numeric <- as.numeric(abcd.data$abuse.binary)
abcd.data$trauma.binary <- factor(abcd.data$trauma>=1,labels = c("No Trauma","Trauma"))
abcd.data$trauma.binary.numeric <- as.numeric(abcd.data$trauma.binary)

median.monitoring = median(na.omit(abcd.data$monitoring))
median.acceptance = median(na.omit(abcd.data$acceptance))
abcd.data$monitoring.binary <- factor(abcd.data$monitoring<20,labels = c("Low Physical Neglect","High Physical Neglect"))
abcd.data$monitoring.binary.numeric <- as.numeric(abcd.data$monitoring.binary)
abcd.data$acceptance.binary <- factor(abcd.data$acceptance<7.5,labels = c("Low Emotional Neglect","High Emotional Neglect"))
abcd.data$acceptance.binary.numeric <- as.numeric(abcd.data$acceptance.binary)
```

## Plot everything
```{r}

data.long <- gather(abcd.data, CogType, CogScore, thompson_PC1,thompson_PC2,thompson_PC3, factor_key=TRUE)
abcd.long <- gather(data.long, ACEsType, ACEsScore, acceptance.binary.numeric,monitoring.binary.numeric,abuse.binary.numeric,trauma.binary.numeric, factor_key=TRUE)

abcd.long.complete <- abcd.long[complete.cases(abcd.long[,c("CogType","CogScore","ACEsType","ACEsScore")]),c("CogType","CogScore","ACEsType","ACEsScore")]

ggplot(data=abcd.long.complete, aes(x=ACEsScore,y=CogScore,fill=interaction(ACEsType,CogType))) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("ACEs") + ylab("Cognition") +coord_cartesian(ylim=c(250,325)) + facet_grid(abcd.long.complete$ACEsType~abcd.long.complete$CogType)

@@

## Run linear mixed effects models!
```{r}

summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+acceptance.binary+monitoring.binary+abuse.binary+trauma.binary,data=abcd.data))

summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+acceptance.binary+monitoring.binary+abuse.binary+trauma.binary,data=abcd.data))

summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+acceptance.binary+monitoring.binary+abuse.binary+trauma.binary,data=abcd.data))

summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+acceptance.binary,data=abcd.data))
summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+monitoring.binary,data=abcd.data))
summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+abuse.binary,data=abcd.data))
summary(lmer(thompson_PC1~interview_age+sex+(1|rel_family_id)+trauma.binary,data=abcd.data))

summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+acceptance.binary,data=abcd.data))
summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+monitoring.binary,data=abcd.data))
summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+abuse.binary,data=abcd.data))
summary(lmer(thompson_PC2~interview_age+sex+(1|rel_family_id)+trauma.binary,data=abcd.data))

summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+acceptance.binary,data=abcd.data))
summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+monitoring.binary,data=abcd.data))
summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+abuse.binary,data=abcd.data))
summary(lmer(thompson_PC3~interview_age+sex+(1|rel_family_id)+trauma.binary,data=abcd.data))

@@

df.summary <- abcd.data %>%
     group_by(ACEs.binary,agesRounded) %>%
     summarise(
         se = sd(thompson_PC1, na.rm = TRUE)/sqrt(length(thompson_PC1)),
         thompson_PC1 = mean(thompson_PC1))
ggplot(data=df.summary, aes(x=ACEs.binary,y=thompson_PC1)) +
  geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("ACEs") + ylab("thompson_PC1") + geom_errorbar(aes(ymin=thompson_PC1-se, ymax=thompson_PC1+se), width=.2,position=position_dodge(.9))+ facet_wrap(~ factor(round(agesRounded)))+coord_cartesian(ylim=c(210,255))

df.summary <- abcd.long %>%
     group_by(ACEsType,ACEsScore,CogType) %>%
     summarise(
         se = sd(CogScore, na.rm = TRUE)/sqrt(length(CogScore)),
         CogScore = mean(na.omit(CogScore)))
ggplot(data=df.summary, aes(x=ACEsScore,y=CogScore,fill=interaction(ACEsType,CogType))) + geom_bar(stat="identity",position="dodge") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black")) +xlab("ACEs") + ylab("Cognition") +coord_cartesian(ylim=c(220,265)) + geom_errorbar(aes(ymin=CogScore-se, ymax=CogScore+se), width=.2,position=position_dodge(.9)) +  facet_grid(df.summary$ACEsType~df.summary$CogType)
```
